---
title: Basic QC and processing of breast cancer single-cell data from Basses et al. (Nature Medicine 2021) 
author: "A. Berenstien; M. Andreatta <massimo.andreatta at unil.ch> and S. Carmona <santiago.carmona at unil.ch>"
output: html_notebook
---

```{r}
renv::activate()
renv::restore()
library(Seurat)
library(dplyr)
library(ggplot2)

```



```{r}
seed = 31416
geo_acc <- "GSE169246"
paper <- "Zhang2021" 
datadir <- paste0("~/Dropbox/FededaBerenstein/data/",paper,"/rawCounts")
out.data.dir <- paste0("~/Dropbox/FededaBerenstein/data/",paper,"/processed")
if(length(list.files(datadir,recursive = T))==0){
  download = T
}else{
  download = F
}

if(download){
    library(GEOquery)
    gse <- getGEO(geo_acc)
    series <- paste0(geo_acc, "_series_matrix.txt.gz")
    system(paste0("mkdir -p ", datadir))
    system(paste0("mkdir -p ", out.data.dir))

    options(timeout=100000)
    #getfiles <- getGEOSuppFiles(geo_acc, baseDir = datadir,fetch_files = F)
    #getGEOSuppFiles(geo_acc, baseDir = datadir,filter_regex = "RNA.counts") #the heaviest one
    getGEOSuppFiles(geo_acc, baseDir = datadir) # Download all files
    
}
```

# Load bcc data
```{r}
rna.seurat.data.file <- file.path(out.data.dir,"Zhang.rna.seurat.rds")
if (!file.exists(rna.seurat.data.file)) {
    system(paste0("mkdir -p ", file.path(datadir,geo_acc,"RNA")))
    rna.files <- list.files(file.path(datadir,geo_acc),pattern = "RNA.")
    for(f in rna.files){
      if(grepl("feature",f)){ link <-"features.tsv.gz"}
      if(grepl("barcode",f)){ link <- "barcodes.tsv.gz"}
      if(grepl("mtx",f)){ link <- "matrix.mtx.gz"}
      file.symlink(from = file.path(datadir,geo_acc,f), to = file.path(datadir,geo_acc,"RNA",link))
    }

    rna.mtx <- Seurat::Read10X(data.dir = file.path(datadir,geo_acc,"RNA"),gene.column=1)
    
    #meatadata files were downloaded from supplementary tables of the original mauscript (xlsx format) and manually converted to tsv
    meta.data.rna <-  read.table(file = file.path(datadir,geo_acc,"rna.metadata.tsv"),sep = "\t",header = T)
    rownames(meta.data.rna) <- meta.data.rna$Cell.barcode
    mm <- match(rna.mtx%>%colnames(),meta.data.rna%>%rownames())
    meta.data.rna <- meta.data.rna[mm,]
    # Create Seurat object and add meta data
    rna.object <- CreateSeuratObject(counts = rna.mtx, project = "Zhang.RNA",
        min.cells = 10, meta.data = meta.data.rna)

    singleCellFeatures::saveRDSMC(rna.object, file = rna.seurat.data.file,threads = 32)

} else {
    rna.object <- readRDS(rna.seurat.data.file)
}
```

# Load ATAC data
```{r}
atac.seurat.data.file <- file.path(out.data.dir,"Zhang.atac.seurat.rds")
load.atac = F
if (!file.exists(atac.seurat.data.file)) {
    system(paste0("mkdir -p ", file.path(datadir,geo_acc,"ATAC")))
    atac.files <- list.files(file.path(datadir,geo_acc),pattern = "ATAC.")
    for(f in atac.files){
      if(grepl("feature",f)) link <- "features.tsv.gz"
      if(grepl("barcode",f)) link <- "barcodes.tsv.gz"
      if(grepl("mtx",f)) link <- "matrix.mtx.gz"

      file.symlink(from = file.path(datadir,geo_acc,f), to = file.path(datadir,geo_acc,"ATAC",link))
    }

    
    atac.mtx <- Seurat::Read10X(data.dir = file.path(datadir,geo_acc,"ATAC"),gene.column=1)
    
    #meatadata files were downloaded from supplementary tables of the original mauscript (xlsx format) and manually converted to tsv
    meta.data.atac <-  read.table(file = file.path(datadir,geo_acc,"atac.metadata.tsv"),sep = "\t",header = T)
    rownames(meta.data.atac) <- meta.data.atac$Cell.barcode
    mm <- match(atac.mtx%>%colnames(),meta.data.atac%>%rownames())
    meta.data.atac <- meta.data.atac[mm,]
    # Create Seurat object and add meta data
    atac.object <- CreateSeuratObject(counts = atac.mtx, project = "Zhang.ATAC",
        min.cells = 10, meta.data = meta.data.atac)

    singleCellFeatures::saveRDSMC(atac.object, file = atac.seurat.data.file,threads = 32)
} else {
  if(load.atac){
    atac.seurat <- readRDS(atac.seurat.data.file)
  }
}

```




Ribosomal and mitochondrial content

```{r, fig.height= 10}
percent.ribo.rna <- PercentageFeatureSet(rna.object, pattern = "^RP[LS]")
percent.mito.rna <- PercentageFeatureSet(rna.object, pattern = "^MT-")

rna.object <- AddMetaData(rna.object, metadata = percent.ribo.rna, col.name = "percent.ribo")
rna.object <- AddMetaData(rna.object, metadata = percent.mito.rna, col.name = "percent.mito")

Idents(rna.object) <- "Patient"
VlnPlot(rna.object, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0)

Idents(rna.object) <- "Group"
VlnPlot(rna.object, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0)

Idents(rna.object) <- "Major.celltype"
VlnPlot(rna.object, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0)

print('Feature RNA')
quantile(rna.object$nFeature_RNA, c(0, 0.01, 0.1, 0.5, 0.9, 0.99))
print('Count RNA')
quantile(rna.object$nCount_RNA, c(0, 0.01, 0.1, 0.5, 0.9, 0.99))
```



Minimal QC to remove extreme outliers
```{r}
dim(rna.object)
rna.object <- subset(rna.object, subset = nFeature_RNA>500 & nFeature_RNA<6000 & 
                       nCount_RNA>600 & nCount_RNA< 21000 &
                       percent.ribo < 50  &
                       percent.mito < 15) # Ya cortaron en 10%
dim(rna.object)
```


### Processing data 
#### (Normalizing, F.selection, scaling, and Reductions)
```{r}
nfeat = 2000
set.seed(seed)
## Process BCC samples
rna.object <- NormalizeData(rna.object, verbose = FALSE);gc()
rna.object <- FindVariableFeatures(rna.object, selection.method = "vst", nfeatures = nfeat, verbose = FALSE);gc()
genes.remove <- unlist(scGate::genes.blacklist.default) #non interesting and/or technical artifact-associated genes
removing <- intersect(rna.object@assays$RNA@var.features, genes.remove)
cat(paste('removing: ',round(100*removing%>%length()/nfeat,2),'% genes\n' ))#, file = con, append =T)
rna.object@assays$RNA@var.features <- setdiff(rna.object@assays$RNA@var.features, genes.remove)
gc()

rna.object <- ScaleData(rna.object); gc()
rna.object <- RunPCA(rna.object, features = rna.object@assays$RNA@var.features, ndims.print = 1:5, nfeatures.print = 5); gc()
rna.object <- RunUMAP(rna.object, reduction = "pca", dims = 1:30, seed.use=seed); gc()

```

## add metadata from Supplementary Table 2
```{r}
cluster.meta.data <-  read.table(file = file.path(datadir,geo_acc,"ts2_zhang2021.tsv"),sep = "\t",header = T)
extended.cluster.meta.data <- merge(rna.object@meta.data[,c("Cell.barcode","Cluster")], cluster.meta.data, by = "Cluster",all.x = T)
mm <- match(rna.object$Cell.barcode,extended.cluster.meta.data$Cell.barcode)
extended.cluster.meta.data <- extended.cluster.meta.data[mm,]
#check
#extended.cluster.meta.data%>%head()
#rna.object@meta.data[1:5,]

# Check excluded (inconsistences ~ 5000 cells)
(extended.cluster.meta.data[extended.cluster.meta.data$Cell.type%>%is.na(),"Cluster"])%>%table()

a <- cluster.meta.data$Cluster%>%unique()
b <- rna.object$Cluster%>%unique()
setdiff(a,b)
setdiff(b,a)

rna.object <- AddMetaData(rna.object,metadata =  as.character(extended.cluster.meta.data$Cell.type), col.name = "Cell.type")
rna.object <- AddMetaData(rna.object,metadata =  extended.cluster.meta.data$Abbreviation,col.name = "Abbreviation")

```

There an inconsistency in table S2, between ILC annotations in shets "Cell clusters" and "Single cell clustering": 
ILC1-SPON2 must be ILC2-SPON2? and ILC2-CNOT2 must be ILC1-CNOT2?
or
ILC1-SPON2 must be ILC1-CNOT2? and ILC2-CNOT2 must be ILC2-SPON2?

Clusters looks consistent with their markers, so the ILCs1/2 labels are wrong?
```{r,fig.width=9,fig.height=3}
exploring.subset <- rna.object%>%subset(Cluster%in%c("t_ILC2-SPON2","t_ILC1-CNOT2"))
a <- DimPlot(exploring.subset,group.by = "Cluster") 
b <- FeaturePlot(exploring.subset,features = c("SPON2","CNOT2"))
ilcs <- rna.object%>%subset(Major.celltype%in%c("ILC cell"))
aux <- ilcs$Cluster%>%strsplit("_")%>%lapply(function(x){x[2]})%>%unlist()%>%strsplit("-")%>%lapply(function(x){x[1]})%>%unlist()
ilcs <- AddMetaData(ilcs,metadata = aux,col.name = "ILC.type")
cc <- DimPlot(ilcs,group.by = "ILC.type") 
d <- DimPlot(ilcs,group.by = "Cluster") + theme(legend.position="bottom") +  theme(legend.key.size = unit(0.2, 'cm'))
a |b
d|cc|a
```



## save processed data for future speed-up 
```{r}
rna.processed.file <- file.path(out.data.dir,'Zhang.rna.processed.rds')
if(!file.exists(rna.processed.file)){
  singleCellFeatures::saveRDSMC(rna.object, rna.processed.file,threads = 32)
}else{
  rna.object <- readRDS(file.path(out.data.dir,'Zhang.rna.processed.rds'))
}
```


```{r,fig.height=5,fig.width=11}
a1 <- DimPlot(rna.object,label = T,repel = T,label.size = 2.5) +   theme(legend.key.size = unit(0.3, 'cm'),aspect.ratio = 1) +labs(subtitle = "Zhang RNA data") 
a2 <- DimPlot(rna.object,label = T,repel = T,label.size = 2.5, group.by = "Abbreviation") + theme(legend.key.size = unit(0.3, 'cm'),aspect.ratio = 1) +labs(subtitle = "Imputed Cell.type") 
a1   + a2

```


```{r,fig.height=5,fig.width=6}
b1 <- DimPlot(rna.object,label = T,repel = T,label.size = 2.5,split.by = "Origin") +   theme(legend.key.size = unit(0.3, 'cm'),aspect.ratio = 1) +labs(subtitle = "Blood/Tumor") 

c1 <- DimPlot(rna.object,label = T,repel = T,label.size = 2.5,split.by = "Group") +   theme(legend.key.size = unit(0.3, 'cm'),aspect.ratio = 1) +labs(subtitle = "Timepoint") 
d1 <- DimPlot(rna.object,label = T,repel = T,label.size = 2.5,split.by = "Tissue") +   theme(legend.key.size = unit(0.3, 'cm'),aspect.ratio = 1) +labs(subtitle = "Tissue") 


#patchwork::wrap_plots(list(a1,a2,b1,c1,d1),design = layout)



```

# Review expression of PD-L1 and PD1 genes per celltype
Aliases
PD-1 ->: PDCD1
PD-L1 -> CD274
```{r,fig.height=5,fig.width=12}
a2 <- DotPlot(rna.object, features = c("PDCD1","CD274")) + RotatedAxis()  +  theme(aspect.ratio = 1) + ggtitle("Major Celltype")
a3 <- DotPlot(rna.object, features = c("PDCD1","CD274"),group.by = "Origin") + RotatedAxis()  +  theme(aspect.ratio = 1) + ggtitle("Origin: Blood/Tumor")

a2 + a3
```


```{r,fig.height=5,fig.width=6}
a5 <- DotPlot(rna.object%>%subset(Origin == "t"), features = c("PDCD1","CD274")) + RotatedAxis()  +  theme(aspect.ratio = 1) + ggtitle("Subset Origin == Tumor")
a5
```

```{r,fig.width=14,fig.height=4}
b2 <- DotPlot(rna.object%>%subset(Group == "Pre-treatment"), features = c("PDCD1","CD274")) + RotatedAxis()  +  theme(aspect.ratio = 1) + ggtitle("Pre-treatment")
c2 <- DotPlot(rna.object%>%subset(Group == "Progression"), features = c("PDCD1","CD274")) + RotatedAxis()  +  theme(aspect.ratio = 1) + ggtitle("Progression")
d2 <- DotPlot(rna.object%>%subset(Group == "Post-treatment"), features = c("PDCD1","CD274")) + RotatedAxis()  +  theme(aspect.ratio = 1) + ggtitle("Post-treatment")
b2 | c2 | d2

```




